Moment Constraints Make Learning Latent-Variable Models Easier

### 2014-01-29 doc/icml2014/background.tex
Fleshed out a bunch for linear algebra in directed
Added words and random mumbo-jumbo

Define $\otimes$ to be the tensor product.
To denote the moments of observed variables, we will use the notation
\begin{align*}
  M_i &\eqdef \Pr(x_i), &
  M_{ij} &\eqdef \Pr(x_i, x_j), &
  M_{ijk} &\eqdef \Pr(x_i, x_j, x_k),
\end{align*}
and so on.
We will view $M_i, M_{ij}, M_{ijk}, \cdots$ as tensors of
  order 1, 2, 3, \ldots in $\Re^D, \Re^{D\times D}, \Re^{D \times
  D \times D}$.
For the moments of hidden variables, we will use analogous notation
\begin{align*}
  P_i &\eqdef \Pr(h_i), &
  P_{ij} &\eqdef \Pr(h_i, h_j), &
  P_{ijk} &\eqdef \Pr(h_i, h_j, h_k).
\end{align*}

@@

Recently, \citet{anandkumar12moments} presented a consistent algorithm
  to learn the parameters of a three-view mixture model
  (\figureref{three-view}) that sidesteps the local optima problem
  altogether. 
In the model, we have three observed variables $x_1, x_2, x_3 \in [d]$
  that are independent conditioned on the latent variable $h \in [k]$. 
  The parameters of this model are the conditional means $O^{(i)} \eqdef
  \E[x_i | h] \in \Re^{D \times K}$ for $i \in \{1,2,3\}$ and the mixture probabilities $\pi
  \in \Delta_{k-1}$.
  The key idea is that the observed moments $M_{12}, M_{23}, M_{13}$ and
  $M_{123}$ admit a factorization in terms of parameters;
\begin{align*}
  M_{ij} &= \sum_h \pi_h O^{(i)}_h \otimes O^{(j)}_h & \forall i,j \in \{1,2,3\} \\
  M_{123} &= \sum_h \pi_h O^{(1)}_h \otimes O^{(2)}_h \otimes O^{(3)}_h.
\end{align*}

\citet{anandkumar13tensor} presented a robust tensor power method, which
  we call $\TensorFactorize$, which converts consistent estimates of the observed moments 
  into consistent estimates of the parameters,
  $\pi, M_1, M_2, M_3$ (up to permutation of the columns), provided
  the following assumption holds;

\begin{assumption}
Assume the conditional moments $M_1, M_2, M_3$ have full column rank
  $k$, and that $\pi \succ 0$.
\end{assumption}

In brief, the algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^T M_{12} W = I_{k \times k}$,
  and uses it to transform $M_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is applied to extract the eigenvectors
  of the whitened $M_{123}$; unwhitening yields the columns of $M_{3}$.
$M_1$ and $M_2$ can then be recovered with some algebra.

>>>>

Define $\otimes$ to be the tensor product.
To denote the moments of observed variables, we will use the notation
\begin{align*}
  \mO_i &\eqdef \Pr(x_i), &
  \mO_{ij} &\eqdef \Pr(x_i, x_j), &
  \mO_{ijk} &\eqdef \Pr(x_i, x_j, x_k),
\end{align*}
and so on.
We will view $\mO_i, \mO_{ij}, \mO_{ijk}, \cdots$ as tensors of
  order 1, 2, 3, \ldots in $\Re^D, \Re^{D\times D}, \Re^{D \times
  D \times D}$.
For the moments of hidden variables, we will use analogous notation
\begin{align*}
  \mH_i &\eqdef \Pr(h_i), &
  \mH_{ij} &\eqdef \Pr(h_i, h_j), &
  \mH_{ijk} &\eqdef \Pr(h_i, h_j, h_k).
\end{align*}

@@

Recently, \citet{anandkumar12moments} presented a consistent algorithm
  to learn the parameters of a three-view mixture model
  (\figureref{three-view}) that sidesteps the local optima problem
  altogether. 
In the model, we have three observed variables $x_1, x_2, x_3 \in [d]$
  that are independent conditioned on the latent variable $h \in [k]$. 
  The parameters of this model are the conditional means $O^{(i)} \eqdef
  \E[x_i | h] \in \Re^{D \times K}$ for $i \in \{1,2,3\}$ and the mixture probabilities $\pi
  \in \Delta_{k-1}$.
  The key idea is that the observed moments $\mO_{12}, \mO_{23}, \mO_{13}$ and
  $\mO_{123}$ admit a factorization in terms of parameters;
\begin{align*}
  \mO_{ij} &= \sum_h \pi_h O^{(i)}_h \otimes O^{(j)}_h & \forall i,j \in \{1,2,3\} \\
  \mO_{123} &= \sum_h \pi_h O^{(1)}_h \otimes O^{(2)}_h \otimes O^{(3)}_h.
\end{align*}

\citet{anandkumar13tensor} presented a robust tensor power method, which
  we call $\TensorFactorize$, which converts consistent estimates of the observed moments 
  into consistent estimates of the parameters,
  $\pi, \mO_1, \mO_2, \mO_3$ (up to permutation of the columns), provided
  the following assumption holds;

\begin{assumption}
\label{asm:full-rank}
Assume the conditional moments $\mO_1, \mO_2, \mO_3$ have full column rank
  $k$, and that $\pi \succ 0$.
\end{assumption}

In brief, the algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^T \mO_{12} W = I_{k \times k}$,
  and uses it to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is applied to extract the eigenvectors
  of the whitened $\mO_{123}$; unwhitening yields the columns of $\mO_{3}$.
$\mO_1$ and $\mO_2$ can then be recovered with some algebra.

### 2014-01-29 doc/icml2014/directed.tex
Updates to paper
Added words and random mumbo-jumbo

We can relate the state of $h_1$ to the moments of $x^a_1$; $M_1
  \eqdef \Pr(x^a_1) = O \pi$. From \assumptionref{full-rank}, $O$ has full
  column rank and thus can be inverted to recover $\pi$; $\pi = \pinv O
  M_1$.

@@

Similarly, the we can write down the moments of $x^a_1, x^a_2$ in terms
of the pairwise marginals $P_{12} \eqdef \Pr(h_1, h_2)$ and solve for
$P_{12}$;
\begin{align*}
  M_{12} &= O P_{12} O^T \\
  P_{12} &= \pinv O M_{12} (\pinv O)^T.
\end{align*}

$T$ can be recovered from the $P_{12}$ by suitable normalization.

@@

Finally, we can estimate $V$ from the third-order moments $M_{234}$ as
follows,
\begin{align*}
  M_{234} &= P_{234}(O, O, O) \\
  P_{234} &= M_{234}(\pinv O, \pinv O, \pinv O).
\end{align*}

@@

%    \COMMENT{Recover observation potentials $O$ using bottlenecks}
      \ENDFOR
%      \COMMENT{\textbf{Step 2:} Recover clique potentials from the piecewise likelihood.}
      \FOR{every clique $\sC = \{h_1, \cdots, h_m\} \in \sG$} 
      \STATE Let $x_\sC = \{x_1, \cdots, x_m\}$.
      \STATE $P_\sC = M_{x_\sC}( \pinv O_{x_1}, \cdots, \pinv O_{x_m} )$.
%      Run expectation-maximization to convergence on the piecewise likelihood \eqref{eqn:piecewise}, over data $\{\vec x_\sC : x \in \sD\}$
      \ENDFOR
  \end{algorithmic}
\end{algorithm}

@@

Finally, we have the following moment condition,
\begin{align*}
  M_{x_\sC} &= P_{\sC}(O_{x_{h_1}}, \cdots, O_{x_{h_m}}) \\
  P_{\sC} &= M_{x_\sC}(\pinv O_{x_{h_1}}, \cdots, \pinv O_{x_{h_m}}).
\end{align*}

@@

When we have multiple observed variables for a hidden variable, it is
  always beneficial to aggregate over them.
Note that $O_h$ will have rank atleast that of $O$; this encodes the
  intuitive fact that we might be able to identify $h$ from multiple
  observed variables even if we can not identify it from a single
  observed variable. 
Suppose $x_\sB(h) = \{x_1, x_2, \cdots, x_l\}$, let $O_h \in
  \Re^{D^l \times K}$ be an
  aggregated observation potential, such that $O_h(x_1, x_2, \cdots,
  x_l) = P(x_1|h) P(x_2|h) \cdots P(x_l|h)$.
Given \assumptionref{full-rank}, $O_h$ have rank $K$.
After step 1, $O_h$ is a known quantity.

\subsection{More examples}

We will now instantiate our algorithm for several examples.

\paragraph{Hidden Markov Model}

\paragraph{Latent Tree Structure}

\paragraph{Directed Grid Model}

>>>>

We can relate the state of $h_1$ to the moments of $x^a_1$; $\mO_1
  \eqdef \Pr(x^a_1) = O \pi$. From \assumptionref{full-rank}, $O$ has full
  column rank and thus can be inverted to recover $\pi$; $\pi = \pinv O
  \mO_1$.

@@

Similarly, the we can write down the moments of $x^a_1, x^a_2$ in terms
of the pairwise marginals $\mH_{12} \eqdef \Pr(h_1, h_2)$ and solve for
$\mH_{12}$;
\begin{align*}
  \mO_{12} &= O \mH_{12} O^T \\
  \mH_{12} &= \pinv O \mO_{12} (\pinv O)^T.
\end{align*}

$T$ can be recovered from the $\mH_{12}$ by suitable normalization.

@@

Finally, we can estimate $V$ from the third-order moments $\mO_{234}$ as
follows,
\begin{align*}
  \mO_{234} &= \mH_{234}(O, O, O) \\
  \mH_{234} &= \mO_{234}(\pinv O, \pinv O, \pinv O).
\end{align*}

@@

%    \COMMENT{Recover observation potentials $O$ using bottlenecks}
      \ENDFOR
%      \COMMENT{\textbf{Step 2:} Recover clique potentials from the piecewise likelihood.}
      \FOR{every clique $\sC = \{h_1, \cdots, h_m\} \in \sG$} 
      \STATE Let $x_\sC = \{x_1, \cdots, x_m\}$.
      \STATE $\mH_\sC = \mO_{x_\sC}( \pinv O_{x_1}, \cdots, \pinv O_{x_m} )$.
%      Run expectation-maximization to convergence on the piecewise likelihood \eqref{eqn:piecewise}, over data $\{\vec x_\sC : x \in \sD\}$
      \ENDFOR
  \end{algorithmic}
\end{algorithm}

@@

Finally, we have the following moment condition,
\begin{align*}
  \mO_{x_\sC} &= \mH_{\sC}(O_{x_{h_1}}, \cdots, O_{x_{h_m}}) \\
  \mH_{\sC} &= \mO_{x_\sC}(\pinv O_{x_{h_1}}, \cdots, \pinv O_{x_{h_m}}).
\end{align*}

@@

When we have multiple observed variables for a hidden variable, it is
  always beneficial to aggregate over them.
Note that $O_h$ will have rank atleast that of $O$; this encodes the
  intuitive fact that we might be able to identify $h$ from multiple
  observed variables even if we can not identify it from a single
  observed variable. 
Suppose $x_\sB(h) = \{x_1, x_2, \cdots, x_l\}$, let $O_h \in
  \Re^{D^l \times K}$ be an
  aggregated observation potential, such that $O_h(x_1, x_2, \cdots,
  x_l) = \mH(x_1|h) \mH(x_2|h) \cdots \mH(x_l|h)$.
Given \assumptionref{full-rank}, $O_h$ have rank $K$.
After step 1, $O_h$ is a known quantity.
